---
title: "R Notebook"
output: html_notebook, pdf_document
---

# Purpose of this notebook
Here I demonstrate how parameters from the Wright-Fisher model with frequency-dependent selection can be inferred from time series data

# Functions to analyse and infer data
Functions to read in data output from Python simulations
```{r}
rm(list=ls())
library(rstan)
library(tidyverse)
library(reticulate)
options(mc.cores=4)

Models <- import_from_path("Models", path = "./Selection/")

# Get first time when each variant appears
fConvertTime <- function(aDF_data){
  lTimeOld <- unique(aDF_data$generation)
  aTimeLookup <- data.frame(old=lTimeOld,new=seq_along(lTimeOld))
  aDF_copy <- aDF_data
  aDF_copy$generation <- aTimeLookup$new[match(aDF_copy$generation,aTimeLookup$old)]
  return(aDF_copy)
}

# Convert variants
fConvertVariants <- function(aDF_data){
  lVarOld <- unique(aDF_data$variant)
  aVarLookup <- data.frame(old=lVarOld,new=seq_along(lVarOld))
  aDF_copy <- aDF_data
  aDF_copy$variant <- aVarLookup$new[match(aDF_copy$variant,aVarLookup$old)]
  aDF_copy <- aDF_copy[order(aDF_copy$generation,aDF_copy$variant),]
  return(aDF_copy)
}

fGetVariantAppearance <- function(aDF_data){
  lVariant <- unique(aDF_data$variant)
  aFirstAppearanceDF <- data.frame(variant=lVariant,appearance=vector(length = length(lVariant)))
  for(i in seq_along(lVariant)){
    aTempDF <- aDF_data[aDF_data$variant==lVariant[i],]
    aFirstAppearanceDF$appearance[i] <- aTempDF$generation[1]
  }
  return(aFirstAppearanceDF)
}

## Get mutant count in each generation
fcountMutants <- function(aGeneration,aDF_data){
  lGens <- aGeneration:(aGeneration + 1)
  aDF_temp <- aDF_data[aDF_data$generation%in%lGens,]
  lVariant_1 <- aDF_temp$variant[aDF_temp$generation==aGeneration]
  lVariant_2 <- aDF_temp$variant[aDF_temp$generation==(aGeneration+1)]
  lMutantVariant <- setdiff(lVariant_2,lVariant_1)
  aMutantDF <- aDF_temp[aDF_temp$variant%in%lMutantVariant,]
  aCount <- sum(aMutantDF$count)
  # print(aCount)
  # stopifnot(all.equal(aMutantDF$count,rep(1,nrow(aMutantDF))))
  return(aCount)
}

fMutantCountsAll <- function(aDF_data){
  aLen <- length(unique(aDF_data$generation))
  lMutantCount <- vector(length=(aLen-1))
  for(i in 1:(aLen-1)){
    lMutantCount[i] <- fcountMutants(i,aDF_data)
  }
  return(lMutantCount)
}

## Create matrix of counts for variants (rows) and generations (cols)
fGetCounts <- function(aDF_short){
  aDF_short <- fConvertVariants(aDF_short)
  K <- length(unique(aDF_short$variant))
  N <- max(aDF_short$generation)
  mCounts <- matrix(nrow = K,ncol = N)
  for(i in 1:N){
    aTempDF <- aDF_short[aDF_short$generation==i,]
    for(j in 1:nrow(aTempDF)){
      mCounts[aTempDF$variant[j],i] <- aTempDF$count[j]
    }
  }
  mCounts[is.na(mCounts)] <- 0
  return(mCounts)
}

## Active variant count
fGetActiveVariantCount <- function(mCounts){
  N <- ncol(mCounts)
  lActiveCount <- vector(length=N)
  for(i in 1:N){
    lTemp <- mCounts[,i]
    lTemp <- which(lTemp!=0,arr.ind = T)
    lActiveCount[i] <- max(lTemp)
  }
  return(lActiveCount) 
}

fPrepareAll <- function(aDF_short){
  lMutantCount <- fMutantCountsAll(aDF_short)
  aFirstAppearances <- fGetVariantAppearance(aDF_short)
  mCounts <- fGetCounts(aDF_short)
  aDF_short <- fConvertTime(aDF_short)
  aDF_short <- fConvertVariants(aDF_short)
  K <- length(unique(aDF_short$variant))
  N <- max(aDF_short$generation)
  lActiveCount <- fGetActiveVariantCount(mCounts)
  data=list(N=N,K=K,firstAppearance=aFirstAppearances$appearance,
                                 mutantCounts=lMutantCount,
                                 counts=mCounts,
                                 activeVariantCount=lActiveCount)
  return(data)
}

```

Inference using Stan
```{r, echo=F, results=F}
aModel <- stan_model('multinomial_freqDep_homogeneous.stan')

initFn <- function(){
  list(s=0,
       delta=0.005)
}
```
# Neutral Wright-Fisher with mutation
Run neutral Wright-Fisher with mutation
```{r}
num_generations <- 1000
pop_size <- 250
mu <- 0.004
selection <- 0
aDF_short <- Models$WrightFisherRun(as.integer(num_generations), as.integer(pop_size), mu, selection, as.integer(pop_size))

aDF_short %>%
  ggplot(aes(x=generation, y=count, fill=as.factor(variant))) +
  geom_area(position = "stack") +
  theme_classic() +
  theme(legend.position = "none") 
```


Fit model to neutral data
```{r, results=F}
data_stan <- fPrepareAll(aDF_short)
fit <- sampling(aModel,data=data_stan,iter=200,chains=4,
                init=initFn)
saveRDS(fit, file = "neutral_example.rds")
```

# Positive frequency dependence Wright-Fisher with mutation
Run positive frequency dependence 
```{r}
selection <- 0.5
aDF_short <- Models$WrightFisherRun(as.integer(num_generations), as.integer(pop_size), mu, selection, as.integer(pop_size))

aDF_short %>%
  ggplot(aes(x=generation, y=count, fill=as.factor(variant))) +
  geom_area(position = "stack") +
  theme_classic() +
  theme(legend.position = "none") 
```

Fit model to positive frequency-dependence data
```{r, results=F}
data_stan <- fPrepareAll(aDF_short)
fit <- sampling(aModel,data=data_stan,iter=200,chains=4,
                init=initFn)
saveRDS(fit, file = "positive_example.rds")
```

# Negative frequency-dependent Wright-Fisher with mutation
Run negative frequency dependence Wright-Fisher
```{r}
selection <- -0.5
aDF_short <- Models$WrightFisherRun(as.integer(num_generations), as.integer(pop_size), mu, selection, as.integer(pop_size))

aDF_short %>%
  ggplot(aes(x=generation, y=count, fill=as.factor(variant))) +
  geom_area(position = "stack") +
  theme_classic() +
  theme(legend.position = "none") 
```


Fit model to negative frequency-dependence data
```{r, results=F}
data_stan <- fPrepareAll(aDF_short)
fit <- sampling(aModel,data=data_stan,iter=200,chains=4,
                init=initFn)
saveRDS(fit, file = "negative_example.rds")
```

# Examining estimated versus actual parameters of Wright-Fisher
## Selection
```{r}
neutral <- readRDS('neutral_example.rds')
positive <- readRDS('positive_example.rds')
negative <- readRDS('negative_example.rds')

alen <- length(rstan::extract(negative)$s)
aTempDF <- tibble(estimated=c(rstan::extract(negative)$s,
                      rstan::extract(neutral)$s,
                      rstan::extract(positive)$s),
                  classification=c(rep("negative", alen),
                           rep("neutral", alen),
                           rep("positive", alen)))
aTempDF$actual <- c(rep(-0.5, alen), rep(0, alen), rep(0.5, alen))
ggplot(aTempDF, aes(x=estimated)) + 
  geom_histogram() +
  geom_vline(aes(xintercept = actual), linetype=2, colour="orange") +
  facet_wrap(~classification) +
  theme_classic()
```

Looks good! The actual vs fitted correspond in each case.

## Mutation rate
```{r}
alen <- length(rstan::extract(negative)$s)
aTempDF <- tibble(estimated=c(rstan::extract(negative)$delta,
                      rstan::extract(neutral)$delta,
                      rstan::extract(positive)$delta),
                  classification=c(rep("negative", alen),
                           rep("neutral", alen),
                           rep("positive", alen)))
aTempDF$actual <- c(rep(0.004, alen), rep(0.004, alen), rep(0.004, alen))
ggplot(aTempDF, aes(x=estimated)) + 
  geom_histogram() +
  geom_vline(aes(xintercept = actual), linetype=2, colour="orange") +
  facet_wrap(~classification) +
  theme_classic() + 
  coord_flip()
```
