---
title: "WF time series Inference"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Purpose of this notebook
Inferring frequency-dependent selection from time series data

# Functions to analyse and infer data
Functions to read in data output from Python simulations
```{r}
rm(list=ls())
library(rstan)
library(tidyverse)
library(reticulate)
options(mc.cores=4)
```
data
```{r}
d<-read.csv("../data/WF_generations4001_5000_allselectionlevels.csv", header=TRUE)
d<-subset(d, S=="0")
```
subset to the number of generations to analyse. 
We will do:
100 generations,<=4100
25 generations, <=4126
10 generations, <=4126
2 generations, <=4003

```{r}
aDF_short<-subset(d, generation<= 4100)
length(unique(aDF_short$generation))
```

```{r}
# Get first time when each variant appears
fConvertTime <- function(aDF_data){
  lTimeOld <- unique(aDF_data$generation)
  aTimeLookup <- data.frame(old=lTimeOld,new=seq_along(lTimeOld))
  aDF_copy <- aDF_data
  aDF_copy$generation <- aTimeLookup$new[match(aDF_copy$generation,aTimeLookup$old)]
  return(aDF_copy)
}

# Convert variants
fConvertVariants <- function(aDF_data){
  lVarOld <- unique(aDF_data$variant)
  aVarLookup <- data.frame(old=lVarOld,new=seq_along(lVarOld))
  aDF_copy <- aDF_data
  aDF_copy$variant <- aVarLookup$new[match(aDF_copy$variant,aVarLookup$old)]
  aDF_copy <- aDF_copy[order(aDF_copy$generation,aDF_copy$variant),]
  return(aDF_copy)
}

fGetVariantAppearance <- function(aDF_data){
  lVariant <- unique(aDF_data$variant)
  aFirstAppearanceDF <- data.frame(variant=lVariant,appearance=vector(length = length(lVariant)))
  for(i in seq_along(lVariant)){
    aTempDF <- aDF_data[aDF_data$variant==lVariant[i],]
    aFirstAppearanceDF$appearance[i] <- aTempDF$generation[1]
  }
  return(aFirstAppearanceDF)
}
```
## Get mutant count in each generation
```{r}
fcountMutants <- function(aGeneration,aDF_data){
  lGens <- aGeneration:(aGeneration + 1)
  aDF_temp <- aDF_data[aDF_data$generation%in%lGens,]
  lVariant_1 <- aDF_temp$variant[aDF_temp$generation==aGeneration]
  lVariant_2 <- aDF_temp$variant[aDF_temp$generation==(aGeneration+1)]
  lMutantVariant <- setdiff(lVariant_2,lVariant_1)
  aMutantDF <- aDF_temp[aDF_temp$variant%in%lMutantVariant,]
  aCount <- sum(aMutantDF$count)
  # print(aCount)
  # stopifnot(all.equal(aMutantDF$count,rep(1,nrow(aMutantDF))))
  return(aCount)
}

fMutantCountsAll <- function(aDF_data){
  aLen <- length(unique(aDF_data$generation))
  lMutantCount <- vector(length=(aLen-1))
  for(i in 1:(aLen-1)){
    lMutantCount[i] <- fcountMutants(i,aDF_data)
  }
  return(lMutantCount)
}

## Create matrix of counts for variants (rows) and generations (cols)
fGetCounts <- function(aDF_short){
  aDF_short <- fConvertVariants(aDF_short)
  K <- length(unique(aDF_short$variant))
  N <- max(aDF_short$generation)
  mCounts <- matrix(nrow = K,ncol = N)
  for(i in 1:N){
    aTempDF <- aDF_short[aDF_short$generation==i,]
    for(j in 1:nrow(aTempDF)){
      mCounts[aTempDF$variant[j],i] <- aTempDF$count[j]
    }
  }
  mCounts[is.na(mCounts)] <- 0
  return(mCounts)
}

## Active variant count
fGetActiveVariantCount <- function(mCounts){
  N <- ncol(mCounts)
  lActiveCount <- vector(length=N)
  for(i in 1:N){
    lTemp <- mCounts[,i]
    lTemp <- which(lTemp!=0,arr.ind = T)
    lActiveCount[i] <- max(lTemp)
  }
  return(lActiveCount) 
}

fPrepareAll <- function(aDF_short){
  lMutantCount <- fMutantCountsAll(aDF_short)
  aFirstAppearances <- fGetVariantAppearance(aDF_short)
  mCounts <- fGetCounts(aDF_short)
  aDF_short <- fConvertTime(aDF_short)
  aDF_short <- fConvertVariants(aDF_short)
  K <- length(unique(aDF_short$variant))
  N <- max(aDF_short$generation)
  lActiveCount <- fGetActiveVariantCount(mCounts)
  data=list(N=N,K=K,firstAppearance=aFirstAppearances$appearance,
                                 mutantCounts=lMutantCount,
                                 counts=mCounts,
                                 activeVariantCount=lActiveCount)
  return(data)
}

```

Inference using Stan
```{r, echo=F, results=F}
aModel <- stan_model('multinomial_freqDep_homogeneous.stan')

initFn <- function(){
  list(s=0,
       delta=0.005)
}
```
Fit model to negative frequency-dependence data
```{r, results=F}
data_stan <- fPrepareAll(aDF_short)
fit <- sampling(aModel,data=data_stan,iter=200,chains=4,
                init=initFn)
saveRDS(fit, file = "WF_s0_g2.rds")
```

# Examining estimated versus actual parameters of Wright-Fisher
## Selection
```{r}
neutral <- readRDS('neutral_example.rds')
positive <- readRDS('positive_example.rds')
negative <- readRDS('negative_example.rds')

alen <- length(rstan::extract(negative)$s)
aTempDF <- tibble(estimated=c(rstan::extract(negative)$s,
                      rstan::extract(neutral)$s,
                      rstan::extract(positive)$s),
                  classification=c(rep("negative", alen),
                           rep("neutral", alen),
                           rep("positive", alen)))
aTempDF$actual <- c(rep(-0.5, alen), rep(0, alen), rep(0.5, alen))
ggplot(aTempDF, aes(x=estimated)) + 
  geom_histogram() +
  geom_vline(aes(xintercept = actual), linetype=2, colour="orange") +
  facet_wrap(~classification) +
  theme_classic()
```

Looks good! The actual vs fitted correspond in each case.

## Mutation rate
```{r}
alen <- length(rstan::extract(negative)$s)
aTempDF <- tibble(estimated=c(rstan::extract(negative)$delta,
                      rstan::extract(neutral)$delta,
                      rstan::extract(positive)$delta),
                  classification=c(rep("negative", alen),
                           rep("neutral", alen),
                           rep("positive", alen)))
aTempDF$actual <- c(rep(0.004, alen), rep(0.004, alen), rep(0.004, alen))
ggplot(aTempDF, aes(x=estimated)) + 
  geom_histogram() +
  geom_vline(aes(xintercept = actual), linetype=2, colour="orange") +
  facet_wrap(~classification) +
  theme_classic() + 
  coord_flip()
```

## Look at simulation results
```{r}
fGetFitted <- function(breaks, s){
  a_max <- length(breaks)
  lLower <- vector(length = (a_max - 1))
  lUpper <- vector(length = (a_max - 1))
  lMedian <- vector(length = (a_max - 1))
  lMu_middle <- vector(length = (a_max - 1))
  lMu_lower <- vector(length = (a_max - 1))
  lMu_upper <- vector(length = (a_max - 1))
  for(i in 2:a_max){
    fit <- readRDS(paste0("../data/inference/WF_s_", s, "_break_", i, ".rds"))
    params <- rstan::extract(fit, 's')[[1]]
    lLower[i] <- quantile(params, 0.025)
    lUpper[i] <- quantile(params, 0.975)
    lMedian[i] <- quantile(params, 0.5)
    params <- rstan::extract(fit, 'delta')[[1]]
    lMu_lower[i] <- quantile(params, 0.025)
    lMu_upper[i] <- quantile(params, 0.975)
    lMu_middle[i] <- quantile(params, 0.5)
  }
  aDF_0 <- tibble(lower=lLower, upper=lUpper, middle=lMedian, breaks=breaks,
                  mu_lower=lMu_lower, mu_middle=lMu_middle, mu_upper=lMu_upper) %>% 
    mutate(s=s)
  return(aDF_0)
}

breaks <- round(readRDS("../data/inference/breaks.rds"))

aDF_0 <- fGetFitted(breaks, 0)
aDF_0.25 <- fGetFitted(breaks, -0.25)
aDF_0.5 <- fGetFitted(breaks, -0.5)
aDF_0.75 <- fGetFitted(breaks, -0.75)
aDF_1 <- fGetFitted(breaks, -1)

```

```{r}
aDF <- bind_rows(aDF_0,
                 aDF_0.25,
                 aDF_0.5,
                 aDF_0.75,
                 aDF_1) %>%
  filter(breaks > 5)
g <- 
  aDF %>% 
  ggplot(aes(x=breaks, fill=as.factor(s))) +
  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.4) +
  geom_line(aes(y=middle, colour=as.factor(s))) +
  scale_x_log10() +
  geom_hline(aes(yintercept=s, colour=as.factor(s)), linetype=2) +
  xlab("Length of series") +
  ylab("Selection") +
  scale_fill_brewer("True\nselection", palette = "Dark2") +
  scale_color_brewer(guide = 'none', palette = "Dark2") +
  theme_classic()
g
ggsave("../outputs/estimated_vs_actual_selection.png", g, dpi = 500, width = 10, height = 6)
saveRDS(aDF, "../data/inference/combined_inference.rds")
```
Mu plot
```{r}
g <- 
  aDF %>% 
  ggplot(aes(x=breaks, fill=as.factor(s))) +
  geom_ribbon(aes(ymin=mu_lower, ymax=mu_upper), alpha=0.4) +
  geom_line(aes(y=mu_middle, colour=as.factor(s))) +
  scale_x_log10() +
  geom_hline(yintercept = 0.001, linetype=2) +
  xlab("Length of series") +
  ylab("Selection") +
  scale_fill_brewer("True\nselection", palette = "Dark2") +
  scale_color_brewer(guide = 'none', palette = "Dark2") +
  theme_classic()
g
ggsave("../outputs/estimated_vs_actual_mu.png", g, dpi = 500, width = 10, height = 6)
```

