---
title: "Generating Moran populations"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

#### Purpose of this notebook
To generate populations using Ben's new selection function. 

### To do
This currently generates one Moran population, with one selection level (0), for 5000000 time steps. But, we have to generate populations with several selection levels, S:

0
-0.01
-0.25
-0.5
-0.75

So this has to be wrapped in a function that steps through the selection levels and run on an HPC.

NOTE: this is a wrapper for python functions which are in another folder and have to be called in "models"

### Functions to analyse and infer data
Functions to read in data output from Python simulations
```{r}
rm(list=ls())
Sys.setenv(RSTUDIO_PANDOC="/anaconda3/bin/pandoc")
library(rstan)
library(tidyverse)
library(reticulate)
options(mc.cores=4)
```
These python modules contain the generating models.
```{r}
Models <- import_from_path("Models", path = "./Selection/")
```
 The reason Moran needs so long is because we run 5000 generations for WF and 5000x1000 to give equivalence for Moran.
 general population parameters. Don't change.
```{r}
num_generations <- 5000000
pop_size <- 1000
mu <- 0.001
```
# Negative frequency-dependent Moran with mutation

Set the selection coefficient. Desired values are: 0, -0.1, -0.2, -0.5, -0.75
```{r}
selection <-0
```
Run the model
```{r}
aDF_short <- Models$MoranRun(as.integer(num_generations), as.integer(pop_size), mu, selection, as.integer(pop_size))
# plot function (optional)
aDF_short %>%
  ggplot(aes(x=generation, y=count, fill=as.factor(variant))) +
  geom_area(position = "stack") +
  theme_classic() +
  theme(legend.position = "none") 
```

```
write to file
```{r}
filename<-paste("Moran", "mu", mu, "S_minus", selection, ".csv", sep="_")
write.csv(aDF_short, "Yule_mu001_s_minus0_10.csv")
```

