---
title: "Generating Yule populations"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

#### Purpose of this notebook
To generate populations using Ben's new selection function. 

### To do
This currently generates one Yule population, with one selection level (0), for 500,000 time steps. But, we have to generate populations with several selection levels, S:

0
-0.01
-0.25
-0.5
-0.75

This has to be wrapped in a function that steps through the selection levels and run on an HPC since 500,000 generations takes a long time. NOTE: 500k was based on Charlie's previous simulations using, i think, a higher mutation rate.  It is possible that we need more in order to generate enough variatns to get a reasnable powerlaw distrbution. this is a matter of look-see. 

NOTE: this is a wrapper for python functions which are in another folder and have to be called in "models"

### Functions to analyse and infer data
Functions to read in data output from Python simulations
```{r}
rm(list=ls())
Sys.setenv(RSTUDIO_PANDOC="/anaconda3/bin/pandoc")
library(rstan)
library(tidyverse)
library(reticulate)
options(mc.cores=4)
```

These python modules contain the generating models.
```{r}
Models <- import_from_path("Models", path = "./Selection/")
```

general population parameters. These do not change (unless you want to test Yule on a shorter run, then change num_generations.)
```{r}
num_generations <- 50000# just for testing, needs to be 500,000 in reality
pop_size <- 1 
mu <- 0.001 
```

Set the selection coefficient. Desired values are: 0, -0.1, -0.25 or -0.5
```{r}
selection <-0
```
Run the model
```{r}
aDF_short <- Models$YuleRun(as.integer(num_generations), as.integer(pop_size), mu, selection, as.integer(pop_size))
```
plot
```{r}
aDF_short %>%
  ggplot(aes(x=generation, y=count, fill=as.factor(variant))) +
  geom_area(position = "stack") +
  theme_classic() +
  theme(legend.position = "none") 
```
write to file
```{r}
filename<-paste("Yule", "mu", mu, "S_minus", selection, ".csv", sep="_")
write.csv(aDF_short, "Yule_mu001_s_minus0_10.csv")
```
